---
title: " Transcriptomic anlaysis on RNASeq for 3 GSCs, 3 Patients"
author: "Ola Hassan"
date: "2/14/2024"
output:
  html_document: default
  pdf_document: default
---

## 3 GBM patients

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r }

#First GSC

GB2_counts <- read.csv("RNAseq GB2 two replicates")
head(GB2_counts)
row.names(GB2_counts) = GB2_counts$X # Assign genes to row names
GB2_counts = GB2_counts[, 2:5]
head(GB2_counts)

GB2_counts["HDAC7",] # validate HDAC7 knockdown in siHHAC7 vs CT
boxplot(GB2_counts) #Quality assesment of the counts




#Second GSC
GB11_counts <- read.csv("RNAseq GB11 two replicates")
head(GB11_counts)
row.names(GB11_counts) = GB11_counts$X
GB11_counts = GB11_counts[, 2:5]
head(GB11_counts)

GB11_counts["HDAC7",] 
boxplot(GB11_counts)

library(dplyr)

library(plyr)

GB24_counts <- read.csv("RNAseq GB24 two replicates")
head(GB24_counts)
row.names(GB24_counts) = GB24_counts$X
GB24_counts = GB24_counts[, 2:5]
head(GB24_counts)

GB24_counts["HDAC7",]

boxplot(GB24_counts)
```


## Loading package DESeq for Differential expression 

```{r }
library(DESeq2)

```

## Buiilding the exp. design and metadata

```{r }

#GB2
exp.des_GB2 <- factor(c("Ctrl", "Ctrl", "Kd", "Kd"))
names(exp.des_GB2) <- colnames(GB2_counts)
type = rep('GSC')

metadata_GB2 = data.frame(colnames(GB2_counts), type, exp.des_GB2,  row.names=NULL)
#write.csv(metadata_GB2, file = 'metadata_GB2.csv')



#GB11 
exp.des_GB11 <- factor(c("Ctrl", "Ctrl", "Kd","Kd"))
names(exp.des_GB11) <- colnames(GB11_counts)
type = rep('GSC')

metadata_GB11= data.frame(colnames(GB11_counts), type, exp.des_GB11,  row.names=NULL)
#write.csv(metadata_GB11, file = 'metadata_GB11.csv')

# GB24

exp.des_GB24 <- factor(c("Ctrl", "Ctrl","Kd","Kd"))
names(exp.des_GB24) <- colnames(GB24_counts)
type = rep('GSC')

metadata_GB24= data.frame(colnames(GB24_counts), type, exp.des_GB24,  row.names=NULL)
#write.csv(metadata_GB24, file = 'metadata_GB24.csv')


```

## Construct DESEQDataSet Object

```{r }


#GB2
DEseq_GB2 <- DESeqDataSetFromMatrix(countData=GB2_counts, 
                              colData= metadata_GB2, 
                              design=~exp.des_GB2)
DEseq_GB2

#GB11
DEseq_GB11 <- DESeqDataSetFromMatrix(countData=GB11_counts, 
                              colData= metadata_GB11, 
                              design=~exp.des_GB11)
DEseq_GB11


#GB24
DEseq_GB24 <- DESeqDataSetFromMatrix(countData=GB24_counts, 
                              colData= metadata_GB24, 
                              design=~exp.des_GB24)
DEseq_GB24


```

## Now run DESEQ function

```{r }

dds_GB2 <- DESeq(DEseq_GB2)
dds_GB11 <- DESeq(DEseq_GB11)
dds_GB24 <- DESeq(DEseq_GB24)

```

## Take a look at the results table

```{r }
##Take a look at the results table



#### The first GSC "GB2"

#GB2 Kd Vs Ctrl
res_GB2 <- results(dds_GB2)
head(res_GB2, tidy=TRUE)



# GB2 Doing a filtering at FDR < 0.01 for Kd Vs Ctrl
res_GB2_Sig_0.05 <- subset(res_GB2, padj < 0.05)

res_GB2_Sig_0.01 <- subset(res_GB2, padj < 0.01)
res_GB2_Sig_0.001 <- subset(res_GB2, padj < 0.001)
res_GB2_Sig_0.0001 <- subset(res_GB2, padj < 0.0001)


nrow(res_GB2_Sig_0.01) # no. of diff expressed genes at FDR, 0.01 = 4963


# GB2 Doing a filtering at FDR < 0.05 for Kd Vs Ctrl
res_GB2_Sig_0.05 <- subset(res_GB2, padj < 0.05)
nrow(res_GB2_Sig_0.05) # no. of diff expressed genes at FDR, 0.01 = 6270


# Save the most diff expressed genes- GB2 "FDR,0.01"
#write.csv(res_GB2_Sig_0.01, file= "Diffexpress_GB2_Sig_0.01_KD~Ctrl.csv", row.names = T)

# Save the most diff expressed genes- GB2 "FDR,0.05"
#write.csv(res_GB2_Sig_0.05, file= "Diffexpress_GB2_Sig_0.05_KD~Ctrl.csv", row.names = T)


# and the downregulated genes by the HDAC7 KD
res_GB2_Sig_0.05_down <- subset(res_GB2_Sig_0.05, log2FoldChange < 0)
nrow(res_GB2_Sig_0.05_down) # no. of diff expressed genes at FDR, 0.01 = 3150
head(res_GB2_Sig_0.05_down)

# Save the down regulated Kd/CTRL- GB2 "FDR,0.05"
#write.csv(res_GB2_Sig_0.05_down, file= "Downregulated_GB2_Sig_0.05_KD~Ctrl.csv", row.names = T)






#### The second GSC "GB11"
#GSCs Kd Vs Ctrl
res_GB11 <- results(dds_GB11)
head(res_GB11, tidy=TRUE)



# GB11 Doing a filtering at FDR < 0.05 for Kd Vs Ctrl
res_GB11_Sig_0.05 <- subset(res_GB11, padj < 0.05)
nrow(res_GB11_Sig_0.05)
# no. of diff expressed genes at FDR, 0.05 = 5284

# GB11 Doing a filtering at FDR < 0.01 for Kd Vs Ctrl
res_GB11_Sig_0.01 <- subset(res_GB11, padj < 0.01)
res_GB11_Sig_0.001 <- subset(res_GB11, padj < 0.001)
res_GB11_Sig_0.0001 <- subset(res_GB11, padj < 0.0001)


nrow(res_GB11_Sig_0.01) # no. of diff expressed genes at FDR, 0.01 = 3907





#and with the strongest upregulation
tail( res_GB11_Sig_0.01[ order( res_GB11_Sig_0.01$log2FoldChange ), ] )

# Save the most diff expressed genes- GB11 "FDR,0.01"
write.csv(res_GB11_Sig_0.01, file= "Diffexpress_GB11_Sig_0.01_KD~Ctrl.csv", row.names = T)

# Save the most diff expressed genes- GB11 "FDR,0.05"
write.csv(res_GB11_Sig_0.05, file= "Diffexpress_GB11_Sig_0.05_KD~Ctrl.csv", row.names = T)


# and the downregulated genes by the HDAC7 KD
res_GB11_Sig_0.05_down <- subset(res_GB11_Sig_0.05, log2FoldChange < 0)
nrow(res_GB11_Sig_0.05_down) # no. of diff expressed genes at FDR, 0.01 = 1870
head(res_GB11_Sig_0.05_down)

# Save the down regulated Kd/CTRL- GB2 "FDR,0.05"
write.csv(res_GB11_Sig_0.05_down, file= "Downregulated_GB11_Sig_0.05_KD~Ctrl.csv", row.names = T)





#### The third GSC "GB24"
#GB24 Kd Vs Ctrl
res_GB24 <- results(dds_GB24)
head(res_GB24, tidy=TRUE)
nrow(res_GB24)


# GB24 Doing a filtering at FDR < 0.05 for Kd Vs Ctrl
res_GB24_Sig_0.05 <- subset(res_GB24, padj < 0.05)
nrow(res_GB24_Sig_0.05)


# GB24 Doing a filtering at FDR < 0.01 for Kd Vs Ctrl
res_GB24_Sig_0.01 <- subset(res_GB24, padj < 0.01)
res_GB24_Sig_0.001 <- subset(res_GB24, padj < 0.001)
res_GB24_Sig_0.0001 <- subset(res_GB24, padj < 0.0001)

nrow(res_GB24_Sig_0.01) # no. of diff expressed genes at FDR, 0.05 = 918 



#and with the strongest upregulation
tail( res_GB24_Sig_0.01[ order( res_GB24_Sig_0.01$log2FoldChange ), ] )

# Save the most diff expressed genes- GB24 "FDR,0.05"
write.csv(res_GB24_Sig_0.05, file= "Diffexpress_GB24_Sig_0.05_KD~Ctrl.csv", row.names = T)

# Save the most diff expressed genes- GB24 "FDR,0.01"
write.csv(res_GB24_Sig_0.01, file= "Diffexpress_GB24_Sig_0.01_KD~Ctrl.csv", row.names = T)


# and the downregulated genes by the HDAC7 KD
res_GB24_Sig_0.05_down <- subset(res_GB24_Sig_0.05, log2FoldChange < 0)
nrow(res_GB24_Sig_0.05_down) # no. of diff expressed genes at FDR, 0.01 = 607
head(res_GB24_Sig_0.05_down)

# Save the down regulated Kd/CTRL- GB2 "FDR,0.05"
write.csv(res_GB24_Sig_0.05_down, file= "Downregulated_GB24_Sig_0.05_KD~Ctrl.csv", row.names = T)




```

## Summary of differential gene expression

```{r }
summary(res_GB2_Sig_0.05)  
summary(res_GB11_Sig_0.05)  
summary(res_GB24_Sig_0.05)  


summary(res_GB2_Sig_0.01)  
summary(res_GB11_Sig_0.01)  
summary(res_GB24_Sig_0.01) 

summary(res_GB2_Sig_0.001)  
summary(res_GB11_Sig_0.001)  
summary(res_GB24_Sig_0.001) 

```

## Diff expressed Sign. genes for FDR <0.05

```{r }
# Diff expressed Sign. genes for GB2
sign_DESeq_GB2_0.05 <- res_GB2[res_GB2$padj<0.05,]# 6270 genes out of 11818
nrow(sign_DESeq_GB2_0.05)
head(sign_DESeq_GB2_0.05)
sign_DESeq_GB2_0.05[order(sign_DESeq_GB2_0.05$padj), ] 
GB2_all_diff_expressed =sign_DESeq_GB2_0.05[order(sign_DESeq_GB2_0.05$log2FoldChange), ] 


# Diff expressed Sign. genes for GB11
sign_DESeq_GB11_0.05 <- res_GB11[res_GB11$padj<0.05,]#  3656
nrow(sign_DESeq_GB11_0.05)
head(sign_DESeq_GB11_0.05)
sign_DESeq_GB11_0.05[order(sign_DESeq_GB11_0.05$padj), ] 
GB11_all_diff_expressed =sign_DESeq_GB11_0.05[order(sign_DESeq_GB11_0.05$log2FoldChange), ] 

# Diff expressed Sign. genes for GB24
sign_DESeq_GB24_0.05 <- res_GB24[res_GB24$padj<0.05,]
nrow(sign_DESeq_GB24_0.05)# 1342 
head(sign_DESeq_GB24_0.05)
sign_DESeq_GB24_0.05[order(sign_DESeq_GB24_0.05$padj), ] 
GB24_all_diff_expressed =sign_DESeq_GB24_0.05[order(sign_DESeq_GB24_0.05$log2FoldChange), ] 


# Matches between both:

GSCs_Matches =intersect(row.names(sign_DESeq_GB2_0.05) , row.names(sign_DESeq_GB11_0.05)) # 2893 genes
GSCs_Matches_all =intersect((GSCs_Matches) , row.names(sign_DESeq_GB24_0.05)) #653 genes




write.csv(GSCs_Matches_all, file= "Diffexpress_3 GSCs_Sig_0.05_KD~Ctrl.csv", row.names = T)



```


## We can use plotCounts fxn to compare the normalized counts between control and KD groups for our top 6 genes 

#GB2

```{r }
 library(tidyverse)

#GB2 
par(mfrow=c(1,3))
plotCounts(dds_GB2, gene="HDAC7", intgroup="exp.des_GB2")
plotCounts(dds_GB24, gene="HDAC7", intgroup="exp.des_GB24")
plotCounts(dds_GB11, gene="HDAC7", intgroup="exp.des_GB11")


HDAC7_GB2 =plotCounts(dds_GB2, gene="CTNNB1", intgroup="exp.des_GB2",returnData = TRUE)
g_HDAC7_GB2 = ggplot(HDAC7_GB2, aes(x= exp.des_GB2, y= count))
g_HDAC7_GB2 + geom_boxplot(aes(fill=exp.des_GB2)) + 
    labs(title="CTNNB1 (GSC1)",
         x="Groups",
         y="Expression (Log normalized counts")


HDAC7_GB24 =plotCounts(dds_GB24, gene="CTNNB1", intgroup="exp.des_GB24",returnData = TRUE)
g_HDAC7_GB24 = ggplot(HDAC7_GB24, aes(x= exp.des_GB24, y= count))
g_HDAC7_GB24 + geom_boxplot(aes(fill=exp.des_GB24)) + 
    labs(title="CTNNB1 (GSC2)",
         x="Groups",
         y="Expression (Log normalized counts")



HDAC7_GB2 =plotCounts(dds_GB2, gene="CHD4", intgroup="exp.des_GB2",returnData = TRUE)
g_HDAC7_GB2 = ggplot(HDAC7_GB2, aes(x= exp.des_GB2, y= count))
g_HDAC7_GB2 + geom_boxplot(aes(fill=exp.des_GB2)) + 
    labs(title="CHD4 (GSC1)",
         x="Groups",
         y="Expression (Log normalized counts")

HDAC7_GB11 =plotCounts(dds_GB11, gene="CHD4", intgroup="exp.des_GB11",returnData = TRUE)
g_HDAC7_GB11 = ggplot(HDAC7_GB11, aes(x= exp.des_GB11, y= count))
g_HDAC7_GB11 + geom_boxplot(aes(fill=exp.des_GB11)) + 
    labs(title="CHD4 (GSC2)",
         x="Groups",
         y="Expression (Log normalized counts")


HDAC7_GB11 =plotCounts(dds_GB11, gene="SMC4", intgroup="exp.des_GB11",returnData = TRUE)
g_HDAC7_GB11 = ggplot(HDAC7_GB11, aes(x= exp.des_GB11, y= count))
g_HDAC7_GB11 + geom_boxplot(aes(fill=exp.des_GB11)) + 
    labs(title="SMC4 (GSC1)",
         x="Groups",
         y="Expression (Log normalized counts")


HDAC7_GB24 =plotCounts(dds_GB24, gene="SMC4", intgroup="exp.des_GB24",returnData = TRUE)
g_HDAC7_GB24 = ggplot(HDAC7_GB24, aes(x= exp.des_GB24, y= count))
g_HDAC7_GB24 + geom_boxplot(aes(fill=exp.des_GB24)) + 
    labs(title="SMC4 (GSC2)",
         x="Groups",
         y="Expression (Log normalized counts")

```

```{r }
 library(tidyverse)

par(mfrow=c(1,3))


HDAC7_GB2 =plotCounts(dds_GB2, gene="HDAC7", intgroup="exp.des_GB2",returnData = TRUE)
g_HDAC7_GB2 = ggplot(HDAC7_GB2, aes(x= exp.des_GB2, y= count))
g_HDAC7_GB2 + geom_boxplot(aes(fill=exp.des_GB2)) + 
    labs(title="HDAC-7 (GB2)",
         x="Groups",
         y="Expression (Log normalized counts")

HDAC7_GB11 =plotCounts(dds_GB11, gene="HDAC7", intgroup="exp.des_GB11",returnData = TRUE)
g_HDAC7_GB11 = ggplot(HDAC7_GB11, aes(x= exp.des_GB11, y= count))
g_HDAC7_GB11 + geom_boxplot(aes(fill=exp.des_GB11)) + 
    labs(title="HDAC-7 (GB11)",
         x="Groups",
         y="Expression (Log normalized counts")

HDAC7_GB24 =plotCounts(dds_GB24, gene="HDAC7", intgroup="exp.des_GB24",returnData = TRUE)
g_HDAC7_GB24 = ggplot(HDAC7_GB24, aes(x= exp.des_GB24, y= count))
g_HDAC7_GB24 + geom_boxplot(aes(fill=exp.des_GB24)) + 
    labs(title="HDAC-7 (GB24)",
         x="Groups",
         y="Expression (Log normalized counts")




```


## PCA
#First we need to transform the raw count data
#vst function will perform variance stabilizing transformation

```{r }
#GB2
vsdata_GB2 <- vst(dds_GB2, blind=FALSE)
plotPCA(vsdata_GB2, intgroup="exp.des_GB2") # using the DESEQ2 plotPCA fxn we can



#GB11
vsdata_GB11 <- vst(dds_GB11, blind=FALSE)
plotPCA(vsdata_GB11, intgroup="exp.des_GB11") # using the DESEQ2 plotPCA fxn we can

#GB24
vsdata_GB24 <- vst(dds_GB24, blind=FALSE)
plotPCA(vsdata_GB24, intgroup="exp.des_GB24") # using the DESEQ2 plotPCA fxn we can


```


## Volcano Plot scales significance P value at y axis and fold change at x axis

```{r }

library(EnhancedVolcano)
library(ggrepel)



EnhancedVolcano(res_GB2_Sig_0.001,
    lab = rownames(res_GB2_Sig_0.001),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlab = bquote(~Log[2]~ 'fold change'),
    title = ' GB2 HDAC7 KD versus CT',
    selectLab = FALSE,
    xlim=c(-3,3),
    ylim=c(0,40),
    FCcutoff = 1,
    pointSize = 1.0,
    labSize = 6.0)







EnhancedVolcano(res_GB2_Sig_0.05,
    lab = rownames(res_GB2_Sig_0.05),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlab = bquote(~Log[2]~ 'fold change'),
    selectLab = c("RAB3GAP2", "ZRF1", "UBE2S", "UBE2T","NOTCH1", "MCM4", "CENPA"),
    title = ' GB2 HDAC7 KD versus CT',
    pCutoff = 10e-1,
    xlim=c(-4,4),
    ylim=c(0,30),
    FCcutoff = 0.5,
    pointSize = 3.0,
    labSize = 6.0)



#GB11

EnhancedVolcano(res_GB11_Sig_0.05,
    lab = rownames(res_GB11_Sig_0.05),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'GB11 HDAC7 KD versus CT',
    pCutoff = 10e-3,
    xlim=c(-2,2),
    ylim=c(0,25),
    FCcutoff = 0.5,
    pointSize = 3.0,
    labSize = 6.0)


EnhancedVolcano(res_GB11_Sig_0.001,
    lab = rownames(res_GB11_Sig_0.001),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlab = bquote(~Log[2]~ 'fold change'),
    title = ' GB2 HDAC7 KD versus CT',
    selectLab = FALSE,
    xlim=c(-3,3),
    ylim=c(0,40),
    FCcutoff = 1,
    pointSize = 1.0,
    labSize = 6.0)

EnhancedVolcano(res_GB11_Sig_0.0001,
    lab = rownames(res_GB11_Sig_0.0001),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlab = bquote(~Log[2]~ 'fold change'),
    title = ' GB2 HDAC7 KD versus CT',
    selectLab = FALSE,
    xlim=c(-3,3),
    ylim=c(0,40),
    FCcutoff = 1,
    pointSize = 1.0,
    labSize = 6.0)

EnhancedVolcano(res_GB11_Sig_0.05,
    lab = rownames(res_GB11_Sig_0.05),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlab = bquote(~Log[2]~ 'fold change'),
    title = ' GB11 HDAC7 KD versus CT',
    cutoffLineType = "blank",
    col = c("grey50", "grey50", "grey50", "red2", "blue" ),
    selectLab = FALSE,
    shadeFill = 'grey',
    xlim=c(-3,3),
    ylim=c(0,40),
    FCcutoff = 2,
    pointSize = 1.0,
    labSize = 6.0)


#GB24

EnhancedVolcano(res_GB24_Sig_0.05,
    lab = rownames(res_GB24_Sig_0.05),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'GB24 HDAC7 KD versus CT',
    pCutoff = 10e-5,
    xlim=c(-2,2),
    ylim=c(0,70),
    FCcutoff = 0.5,
    pointSize = 3.0,
    labSize = 6.0)

EnhancedVolcano(res_GB24_Sig_0.01,
    lab = rownames(res_GB24_Sig_0.01),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'GB24 HDAC7 KD versus CT',
    pCutoff = 10e-5,
    xlim=c(-2,2),
    cutoffLineType = "blank",
    selectLab = FALSE,
    ylim=c(0,30),
    FCcutoff = 0.5,
    pointSize = 1.0,
    labSize = 6.0)


EnhancedVolcano(res_GB24_Sig_0.001,
    lab = rownames(res_GB24_Sig_0.001),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'GB24 HDAC7 KD versus CT',
    pCutoff = 10e-5,
    xlim=c(-2,2),
    selectLab = FALSE,
    ylim=c(0,30),
    FCcutoff = 0.5,
    pointSize = 1.0,
    labSize = 6.0)

EnhancedVolcano(res_GB24_Sig_0.05,
    lab = rownames(res_GB24_Sig_0.05),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlab = bquote(~Log[2]~ 'fold change'),
    title = ' GB11 HDAC7 KD versus CT',
    cutoffLineType = "blank",
    col = c("grey50", "grey50", "grey50", "red2", "blue"),
    selectLab = FALSE,
    shadeFill = 'grey',
    xlim=c(-3,3),
    ylim=c(0,40),
    FCcutoff = 2,
    pointSize = 1.0,
    labSize = 6.0)

```

## MA plot : MA plot describes the differential expression by plotting log transformed expression values with mean average values, while volcano plot scales significance value at y axis and fold change at x axis

```{r }
par(mfrow=c(1,3))


#Red points have adjusted p value < 0.1. We see that in general genes that are highly expressed, or have a high fold change, are DE.
plotMA(res_GB2_Sig_0.01, ylim=c(-7,7), main="MA plot_GB2")
plotMA(GB11_all_diff_expressed, ylim=c(-7,7), main="MA plot_GB11")
plotMA(GB24_all_diff_expressed, ylim=c(-7,7), main="MA plot_GB24")

```



## Load Pheatmap package

```{r }
library("pheatmap")
library("RColorBrewer") # Load a package giving more colors

```

## Do the rlog transformation:


```{r }
rld_GB2 = rlogTransformation(dds_GB2)
rld_GB11 = rlogTransformation(dds_GB11)
rld_GB24 = rlogTransformation(dds_GB24)


##Clustering the sample-to-sample distances
##Pheatmap

#GB2
distsRL_GB2 <- dist(t(assay(rld_GB2))) # Calculate distances using transformed (and normalized) counts
mat_GB2 <- as.matrix(distsRL_GB2) # convert to matrix
rownames(mat_GB2) <- colnames(mat_GB2) <- with(colData(dds_GB2), paste(exp.des_GB2)) # set rownames in the matrix
colnames(mat_GB2) = NULL # remove column names
par(ps = 9, cex = 0.8, cex.main = 1)
colors <- colorRampPalette( rev(brewer.pal(9, "Reds")) )(255) # set colors
pheatmap(mat_GB2,
         clustering_distance_rows=distsRL_GB2,
         clustering_distance_cols=distsRL_GB2,
         col=colors)


#GB11
distsRL_GB11 <- dist(t(assay(rld_GB11))) # Calculate distances using transformed (and normalized) counts
mat_GB11 <- as.matrix(distsRL_GB11) # convert to matrix
rownames(mat_GB11) <- colnames(mat_GB11) <- with(colData(dds_GB11), paste(exp.des_GB11)) # set rownames in the matrix
colnames(mat_GB11) = NULL # remove column names
par(ps = 9, cex = 0.8, cex.main = 1)
colors <- colorRampPalette( rev(brewer.pal(9, "Reds")) )(255) # set colors
pheatmap(mat_GB11,
         clustering_distance_rows=distsRL_GB11,
         clustering_distance_cols=distsRL_GB11,
         col=colors)


#GB24
distsRL_GB24 <- dist(t(assay(rld_GB24))) # Calculate distances using transformed (and normalized) counts
mat_GB24 <- as.matrix(distsRL_GB24) # convert to matrix
rownames(mat_GB24) <- colnames(mat_GB24) <- with(colData(dds_GB24), paste(exp.des_GB24)) # set rownames in the matrix
colnames(mat_GB24) = NULL # remove column names
par(ps = 9, cex = 0.8, cex.main = 1)
colors <- colorRampPalette( rev(brewer.pal(9, "Reds")) )(255) # set colors
pheatmap(mat_GB24,
         clustering_distance_rows=distsRL_GB24,
         clustering_distance_cols=distsRL_GB24,
         col=colors)

```

## Heatmap of the most significantly differentially expressed genes using Pheatmap library


```{r }
#GB2




Top_500sig_GB2 = mat_GB2 = assay(rld_GB2)[ head(order(res_GB2_Sig_0.05$padj),500), ] # select the top 500 genes withthe lowest padj

Top_500sig_GB2 = mat_GB2 = assay(rld_GB2)[ head(order(res_GB2_Sig_0.05$log2FoldChange),20), ] # select the top 500 genes withthe lowest padj


mat_GB2 = mat_GB2 - rowMeans(mat_GB2) # Subtract the row means from each value
# To make the plot nicer:
df_GB2 = as.data.frame(colData(rld_GB2)[,c("exp.des_GB2")]) # Create a dataframe with a column of the exp.design
colnames(df_GB2) = "exp.des_GB2"   # Rename the column header
rownames(df_GB2) = colnames(mat_GB2)  # add rownames

# and plot the actual heatmap
#pheatmap(mat_GB2, annotation_col=df_GB2,  fontsize_row = 6, pheatmap(mat_GB24, annotation_col=df_GB24, fontsize_row = 6, color = colorRampPalette(c("purple3", "white", "green"))(50)))




#GB11

Top_500sig_GB11 = mat_GB11 = assay(rld_GB11)[ head(order(res_GB11_Sig_0.05$padj),20), ] # select the top 500 genes withthe lowest padj

Top_1000sig_GB11 = mat_GB11 = assay(rld_GB11)[ head(order(res_GB11_Sig_0.01$log2FoldChange),20), ] # select the top 1000 genes withthe lowest padj


mat_GB11 = mat_GB11 - rowMeans(mat_GB11) # Subtract the row means from each value
# To make the plot nicer:
df_GB11 = as.data.frame(colData(rld_GB11)[,c("exp.des_GB11")]) # Create a dataframe with a column of the exp.design
colnames(df_GB11) = "exp.des_GB11"   # Rename the column header
rownames(df_GB11) = colnames(mat_GB11)  # add rownames

# and plot the actual heatmap
pheatmap(mat_GB11, annotation_col=df_GB11,  fontsize_row = 6, color = colorRampPalette(c("purple3", "white", "green"))(50))

#GB24

Top_500sig_GB24 = mat_GB24 = assay(dds_GB24)[ head(order(res_GB24_Sig_0.05$log2FoldChange),30), ] # select the top 500 genes withthe lowest padj

Top_1000sig_GB24 = mat_GB24 = assay(rld_GB24)[ head(order(GB24_all_diff_expressed$log2FoldChange),30), ] # select the top 500 genes withthe lowest padj

mat_GB24 = mat_GB24 - rowMeans(mat_GB24) # Subtract the row means from each value
# To make the plot nicer:
df_GB24 = as.data.frame(colData(rld_GB24)[,c("exp.des_GB24")]) # Create a dataframe with a column of the exp.design
colnames(df_GB24) = "exp.des_GB24" # Rename the column header
rownames(df_GB24) = colnames(mat_GB24)  # add rownames

# and plot the actual heatmap
pheatmap(mat_GB24, annotation_col=df_GB24, fontsize_row = 6, color = colorRampPalette(c("purple3", "white", "green"))(50))




```




```{r }
library(clusterProfiler)
library(org.Hs.eg.db)

# GB2


sign_DESeq_GB2_0.001 <- res_GB2[res_GB2$padj<0.001,]# 6270 genes out of 11818
nrow(sign_DESeq_GB2_0.001)
head(sign_DESeq_GB2_0.001)
sign_DESeq_GB2_0.001[order(sign_DESeq_GB2_0.001$padj), ] 
GB2_all_diff_expressed =sign_DESeq_GB2_0.001[order(sign_DESeq_GB2_0.001$log2FoldChange), ] 

genes_0.01 = sign_DESeq_GB2_0.001[order(sign_DESeq_GB2_0.001$log2FoldChange), ] 

gene_df = data.frame(rownames(genes_0.01), genes_0.01$log2FoldChange)

gene_list = data.frame(rownames(genes_0.01))


ego <- enrichGO(gene         = rownames(genes_0.01),
                OrgDb         = org.Hs.eg.db,
                keyType       = 'SYMBOL',
                ont           = 'BP',
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05)
head(summary(ego))

dotplot(ego, showCategory=10)
barplot(ego, showCategory=20, descending = TRUE, font.size = 18 )



goplot(ego, 10)

#GB11
genes_GB11 = sign_DESeq_GB11_0.05[order(sign_DESeq_GB11_0.05$log2FoldChange), ] 

gene_df_GB11 = data.frame(rownames(genes_GB11), genes_GB11$log2FoldChange)

gene_list_GB11 = data.frame(rownames(genes_GB11))


ego_GB11 <- enrichGO(gene         = rownames(genes_GB11),
                OrgDb         = org.Hs.eg.db,
                keyType       = 'SYMBOL',
                ont           = 'BP',
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05)
head(summary(ego_GB11))

dotplot(ego_GB11, font.size = 10,showCategory=30)

goplot(ego_GB11, font.size = 8)


#GB24
genes_GB24 = sign_DESeq_GB24_0.05[order(sign_DESeq_GB24_0.05$log2FoldChange), ] 

gene_df_GB24 = data.frame(rownames(genes_GB24), genes_GB24$log2FoldChange)

gene_list_GB24 = data.frame(rownames(genes_GB24))


ego_GB24 <- enrichGO(gene         = rownames(genes_GB24),
                OrgDb         = org.Hs.eg.db,
                keyType       = 'SYMBOL',
                ont           = 'BP',
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05)
head(summary(ego_GB24))

dotplot(ego_GB24, font.size = 10,
 showCategory=30)

goplot(ego_GB24, font.size = 8, showCategory = 5)



```


Functional enrichment analysis by GSEA
```{r }

library(clusterProfiler)
library(enrichplot)
library(ggplot2)

# SET THE DESIRED ORGANISM HERE
library(org.Hs.eg.db)

HA = org.Hs.eg.db
```


Functional enrichment analysis by GSEA
```{r }

# reading in data from deseq2
original_gene_list_GB2 <- res_GB2_Sig_0.001$log2FoldChange


# name the vector
names(original_gene_list_GB2) <- row.names(res_GB2_Sig_0.001)  
#write.csv(original_gene_list_GB2, file= "GB2_Sig_0.001.csv", row.names = T)


# sort the list in decreasing order (required for clusterProfiler)
gene_list_GB2 = sort(original_gene_list_GB2, decreasing = TRUE)

gse_GB2 <- gseGO(geneList=gene_list_GB2, 
             ont ="ALL",
             keyType = "SYMBOL", 
             nPerm = 1000, 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.01,
             verbose = TRUE, 
             OrgDb = HA, 
             pAdjustMethod = "none")

require(DOSE)
emapplot(gse_GB2, showCategory = 10)




#ego <- enrichGO(gene         = rownames(genes_GB2),
               # OrgDb         = org.Hs.eg.db,
              #  keyType       = 'SYMBOL',
              #  ont           = 'BP',
             #   pAdjustMethod = "BH",
              #  pvalueCutoff  = 0.05,
              #  qvalueCutoff  = 0.05)
#head(summary(ego))

#dotplot(ego, showCategory=10)
#barplot(ego, showCategory=10)





# reading in data from deseq2
#GB11
# reading in data from deseq2
original_gene_list_GB11 <- res_GB11_Sig_0.01$log2FoldChange

# name the vector
names(original_gene_list_GB11) <- row.names(res_GB11_Sig_0.01)  

# sort the list in decreasing order (required for clusterProfiler)
gene_list_GB11 = sort(original_gene_list_GB11, decreasing = TRUE)

gse_GB11 <- gseGO(geneList=gene_list_GB11, 
             ont ="ALL",
             keyType = "SYMBOL", 
             nPerm = 1000, 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.01,
             verbose = TRUE, 
             OrgDb = HA, 
             pAdjustMethod = "none")
emapplot(gse_GB11, showCategory = 10, color = "p.adjust",)


#GB24
# reading in data from deseq2
original_gene_list_GB24 <- res_GB24_Sig_0.01$log2FoldChange

# name the vector
names(original_gene_list_GB24) <- row.names(res_GB24_Sig_0.01)  

# sort the list in decreasing order (required for clusterProfiler)
gene_list_GB24 = sort(original_gene_list_GB24, decreasing = TRUE)

gse_GB24 <- gseGO(geneList=gene_list_GB24, 
             ont ="ALL",
             keyType = "SYMBOL", 
             nPerm = 1000, 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.01,
             verbose = TRUE, 
             OrgDb = HA, 
             pAdjustMethod = "none")

emapplot(gse_GB24, showCategory = 12)

```

```{r }
#GB2

GB2_counts["FGF2",]
GB2_counts["ICMT" ,]
GB2_counts["ORC1",]

GB2_counts["H2AFX",]
GB2_counts["H2AFZ",]

GB2_counts["PCBP1",]
GB2_counts["RANGAP1",]
GB2_counts["AURKA",]
GB2_counts["TUBG1",]
GB2_counts["SLC7A5",]
GB2_counts["CDC45",]

GB2_counts["BCLAF1",]
GB2_counts["CEBPZ",]
GB2_counts["CCT7",]
GB2_counts["SEC24A",]
GB2_counts["TPX2",]
GB2_counts["SF3B1",]
GB2_counts["ORC6",]
GB2_counts["RAB3GAP2",]
GB2_counts["FZR1",]
GB2_counts["UBE2S",]
GB2_counts["UBE2T",]
GB2_counts["TNFRSF12A",]
GB2_counts["CCNB1",]
GB2_counts["MCM3",]
GB2_counts["MCM5",]
GB2_counts["MCM7",]
GB2_counts["MCM4",]
GB2_counts["MCM6",]
GB2_counts["SMC4",]
GB2_counts["SMC2",]
GB2_counts["EIF2B3",]
GB2_counts["NOTCH1",]
GB2_counts["CDC6",]
GB2_counts["DNMT1",]
GB2_counts["MYBL2",]
GB2_counts["BRAF",]
GB2_counts["E4F1",]
GB2_counts["OLIG2",]
GB2_counts["KDM7A",]
GB2_counts["CHD4",]
GB2_counts["KDM6B",]
GB2_counts["VGF",]
GB2_counts["RAF1",]
GB2_counts["YAF2",]
GB2_counts["JMJD6",]
GB2_counts["MYC",]
GB2_counts["KMT2B",]
GB2_counts["E2F2",]
GB2_counts["SETD1A",]
GB2_counts["WNT4",]
GB2_counts["FGF18",]




#GB24
GB24_counts["FGF18",]

GB24_counts["FGF2",]
GB24_counts["ICMT" ,]
GB24_counts["ORC1",]
GB24_counts["H2AFX",]
GB24_counts["H2AFZ",]
GB24_counts["RANGAP1",]
GB24_counts["CDC45",]
GB24_counts["BCLAF1",]
GB24_counts["CEBPZ",]
GB24_counts["ORC6",]
GB24_counts["RAB3GAP2",]
GB24_counts["UBE2S",]
GB24_counts["UBE2T",]
GB24_counts["TNFRSF12A",]
GB24_counts["CCNB1",]
GB24_counts["MCM3",]
GB24_counts["MCM5",]
GB24_counts["MCM7",]
GB24_counts["MCM4",]
GB24_counts["MCM6",]
GB24_counts["SMC4",]
GB24_counts["SMC2",]
GB24_counts["EIF2B3",]
GB24_counts["NOTCH1",]
GB24_counts["CDC6",]
GB24_counts["DNMT1",]
GB24_counts["MYBL2",]
GB24_counts["BRAF",]
GB24_counts["E4F1",]
GB24_counts["OLIG2",]
GB24_counts["KDM7A",]
GB24_counts["CHD4",]
GB24_counts["KDM6B",]
GB24_counts["VGF",]
GB24_counts["RAF1",]
GB24_counts["YAF2",]
GB24_counts["JMJD6",]
GB24_counts["MYC",]
GB24_counts["KMT2B",]
GB24_counts["E2F2",]
GB24_counts["SETD1A",]
GB24_counts["WNT4",]
GB24_counts["CDC45",]


#GB11
GB11_counts["FGF18",]
GB11_counts["ICMT" ,]
GB11_counts["ORC1",]
GB11_counts["H2AFX",]
GB11_counts["H2AFZ",]
GB11_counts["RANGAP1",]
GB11_counts["CDC45",]
GB11_counts["BCLAF1",]
GB11_counts["CEBPZ",]
GB11_counts["ORC6",]
GB11_counts["RAB3GAP2",]
GB11_counts["UBE2S",]
GB11_counts["UBE2T",]
GB11_counts["TNFRSF12A",]
GB11_counts["CCNB1",]
GB11_counts["MCM3",]
GB11_counts["MCM5",]
GB11_counts["MCM7",]
GB11_counts["MCM4",]
GB11_counts["MCM6",]
GB11_counts["SMC4",]
GB11_counts["SMC2",]
GB11_counts["EIF2B3",]
GB11_counts["NOTCH1",]
GB11_counts["CDC6",]
GB11_counts["DNMT1",]
GB11_counts["MYBL2",]
GB11_counts["BRAF",]
GB11_counts["E4F1",]
GB11_counts["OLIG2",]
GB11_counts["KDM7A",]
GB11_counts["CHD4",]
GB11_counts["KDM6B",]
GB11_counts["VGF",]
GB11_counts["RAF1",]
GB11_counts["YAF2",]
GB11_counts["JMJD6",]
GB11_counts["MYC",]
GB11_counts["KMT2B",]
GB11_counts["E2F2",]
GB11_counts["SETD1A",]
GB11_counts["WNT4",]
GB11_counts["CDC45",]
GB11_counts["SOCS1",]

GB11_counts

```



## Transcrption factor enrichment analysis:
**We put the genes that were downregulated by the HDAC7 knockdown from the 3GSCs on the ChEA3 webtool, https://maayanlab.cloud/chea3/**

## Stemness enrichment and phenotype analysis:
**We put the genes that were downregulated by the HDAC7 knockdown from the 3 GSCs on the StemChecker webtool, http://stemchecker.sysbiolab.eu**

## Stemness gene signature analysis:
**We put the genes that were downregulated by the HDAC7 knockdown from the 3 GSCs on the GSEA software **
## References:

 *https://bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html*
 *http://bioconnector.github.io/bims8382/r-rnaseq-airway.html*
 *http://stemchecker.sysbiolab.eu/*

 
